When Proximity Isn't Enough: Network Segregation and SES Bias in Referrals

### 2025-08-06 writing/manuscript.tex
v1.1.7
Chapter 3 to-do: finish re-read and send

\title{Class differences in social networks: Evidence from a referral experiment}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz\footremember{liser}{Luxembourg Institute of Socio-Economic Research}, Ernesto Reuben\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser}, Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether differences in high-SES connection shares are exacerbated by SES biases in referral selection. We conduct a lab-in-the-field experiment with 734 Colombian university students making incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed \$25 bonus for referral recipients, we find no effect of the treatment on referral choices: Referrals consistently go to high-performing peers with whom they take many courses together. We find that while low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no bias toward other groups. Network segregation, driven by students selecting into programs based on their SES, limits cross-SES referral opportunities even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\pagebreak
\section{Introduction}
Equally qualified individuals in terms of productivity face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectivity, defined as the share of high-SES connections among low-SES individuals, is the most important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Research on economic connectivity has focused on two distinct mechanisms that shape cross-SES connections: network composition (who you have the chance to meet inside an institutional environment) versus individual preference (who you choose to connect with among those available). The prevailing hypothesis emerging from the seminal work of \citet{chetty2022n} is that increasing exposure to high-SES individuals under favorable intergroup contact conditions will lead low-SES individuals to connect with them at higher rates. Universities, in this regard, represent a particularly promising setting as they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, credible evidence on biases in individual preferences to connect across SES groups remains limited. One important reason for this gap is the challenge of creating controlled environments that isolate SES biases while accounting for natural variations in social network compositions.

We overcome this challenge through a lab-in-the-field experiment at a Colombian university. Focusing on the role of SES in referral selection, we studied whether individuals who were asked to refer a peer tended to refer a same-SES candidate. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam, and to incentivize performance-based referral selection, participants earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals while SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course the referrer had taken with all other undergraduate students at the university (more than 4,500 individuals). It allowed us to observe both characteristics of every potential referral candidate, and the number of courses taken together as a measure of the intensity of interaction between the referrer and the candidate. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. We designed the \textbf{Bonus} condition to make SES biases in referral selection even more salient. The fixed bonus created incentives to refer peers with high connection intensity (many courses taken together) even if they performed less well, potentially amplifying SES bias since connection intensity and SES background are correlated.

We find that referrals consistently go to higher-performing peers with strong connection intensity (14 vs. 4 courses taken together), regardless of the incentive conditions and the exam area. Pooling across conditions and exam areas, we find that SES bias in referrals is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no bias toward other groups.

With 93\% of referrals going to peers within the same academic program with whom referrers have taken many courses together, we find that network composition rather than SES biases explain most observed referral patterns. At the connection intensity where referrals typically occur (median 12 courses together), network segregation becomes stark: low-SES networks contain 44.5\% low-SES peers versus 35\% university-wide (27\% increase), while high-SES networks contain only 15.7\% low-SES peers (55\% decrease from the university average). This segregation means that even without any bias against low-SES peers, high-SES referrers rarely encounter low-SES candidates in their practical choice sets.

Looking for potential mechanisms driving the segregation in enrollment networks, we identify program selection as key. Program fees at the partner university are fixed on a cost basis, and with less than 5\% of undergraduates qualifying for scholarships. The consequence of these institutional policies is that SES groups end up sorting into programs on the basis of their costs, where some programs cost up to six times more on a yearly basis. To sum, even though low-SES are exposed to higher-than-population shares of high-SES students, and high-SES are not biased toward other SES groups, meaningful interaction opportunities are limited due to institutional and organizational factors.

Our findings should be interpreted with some scope conditions. First, our referrals have no direct job consequences, and referring under anonymity presents a lower stake environment with no potential reputational concerns. Nevertheless, we replicate typical findings from earlier referral experiments where performance-based incentives brings in qualified candidates (e.g., \citet{witte2021workers,beaman_who_2012}).

Second, our enrollment networks capture classroom-based interactions and a measure for their intensity rather than broader social networks and close friendships. This approach offers advantages over self-reported friendship network elicitation methods, which suffers from censoring due to size limitations \citep{griffith2022jle}, or social media networks, which may not overlap with actual interactions at the university. We find that courses taken together predicts referral selection beyond program affiliation, suggesting it does capture meaningful variation in connection intensity.

Finally, our setting examines SES bias within a single institution where cross-SES contact is possible, but where the networks of different SES groups are separated due to program selection. The generalizability to contexts with different institutional structures remains an open question for future research.

@@

A subset of the literature focuses on SES-based differences in social capital and network formation \citep{chetty2022na,engzell2025ajos,michelman2022tqjoe}, with connection intensity \citep{gee2017jole,kramarz2014res,wang2013aejae,sterling2014ms} and homophily \citep{mcpherson2001birds,currarini2009e,jackson2022,bolte2024jle,montgomery_social_1991} driving differences across groups. Based on the pioneering work of \citet{currarini2010pnas}, we contribute by identifying two different types of homophily, and separate whether differential outcomes stem from network composition (who you know) versus taste-based biases (who you choose to interact with). Our findings suggest that rather than focusing on taste-based biases,  implementing mixed-program courses to increase across-SES connection intensity should be a clear policy goal in order to reduce SES-based network segregation (see e.g., for a similar institutional intervention \citet{rohrer2021po}).

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} and their consequences. Performance-based incentives bring in qualified candidates otherwise not identified by demographics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by a simple referral framework where we replicate the earlier findings and causally evaluate the effect of a sizeable monetary bonus for the referral candidate.

The remainder of the paper is organized as follows. Section \ref{BS} begins with the background and setting in Colombia. In Section \ref{ES} we present the empirical strategy and in Section \ref{Design} we present the design of the experiment. In Section \ref{SID} we describe the experimental sample, incentives and the procedure. Section \ref{Results} discusses the results of the experiment and Section \ref{MandRC} discusses potential mechanisms and robustness checks. Section \ref{Conclusion} concludes. The Appendix presents additional tables and figures as well as the experiment instructions.

@@

\subsection{Inequality and SES in Colombia}
Our experiment took place in Colombia, a country that consistently ranks highly in terms of economic inequality. The richest decile of Colombians earn 50 times more than the poorest decile \citep{barretoherrera2024regional,comisioneconomicaparaamericalatinayel2023}. This economic disparity creates profound differences in outcomes across SES groups in terms of education, geographic residence, language, manners, and social networks \citep{garciavillegas2021dimension, angulo2012movilidad,garcia2015loteria}. While these patterns are not atypical and exist elsewhere, Colombia's pronounced inequality makes economic, educational, and cultural differences across SES particularly visible.

We rely on Colombia's established estrato classification system to measure SES in our study. In 1994, Colombia introduced a nationwide system that divides the population into six strata based on  ``similar social and economic characteristics'' \citep[p. 102]{hudson_colombia_2010}. Designed for utility subsidies from higher strata to support lower strata, the system aligns with and reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}. It is widely used by policymakers and in official statistics \citep{fergusson2021desigualdad}. Using the estrato system, we categorize students in strata 1-2 as low-SES, strata 3-4 as middle-SES, and strata 5-6 as high-SES.

@@

\begin{figure}[H]
  \centering
  \caption{Income, performance, and university choice in Colombia}\label{school_quality_income}
  \includegraphics[width=0.6\linewidth]{../figures/school_quality_income.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure shows the average score national university entry exam by monthly family income and type of higher education institution. With average student scores in the 65-70 band (see Appendix Figures \ref{math_hist} and \ref{reading_hist}), the private university where we conducted this experiment caters to low-, middle- and high-SES students. Figure reproduced from \citet{fergusson2021distincion}.
  \end{tablenotes}
\end{figure}

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 and 7 courses per semester, and programs last anywhere between 4 and 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters.

To map social connections at the university, we construct enrollment networks using administrative data. For each participant, we identify all other undergraduate students with whom they have taken at least one course and create their individual network of university connections. The size of this network depends on how many students a participant has encountered through coursework, while the connection intensity is measured by the number of courses taken together. This approach provides a complete picture of each participant's social environment at the university, including detailed characteristics (i.e., SES, academic program, performance) for both the participant and every person in their network.

@@

where $Y_{ij} = 1$ if referrer $i$ selects candidate $j$, and 0 otherwise. We set middle-SES as the base category, so $\beta_1$ is the log-odds estimate for referring low- and high-SES candidates relative to middle-SES. $X_{ij}$ includes the remaining characteristics of referral candidates in the enrollment network that improve model fit such as  entry exams scores and the number of courses taken together with the referrer. These continuous variables are standardized using means and standard deviations calculated by first computing network-level statistics for each referrer, then averaging across all 734 networks.\footnote{Each referral candidate's entry exam score and the number of courses they have taken with the referrer is standardized using these sample-level statistics. The standardization formula is $z_{i} = (x_{i} - \bar{X}) / \sigma$, where $\bar{X}$ and $\sigma$ are the average mean and standard deviation across participant networks for the measure.} The individual fixed effects $\alpha_i$ control for all referrer-specific factors that might influence both network formation and referral decisions. Because we observe two referrals from each referrer, we cluster standard errors at the referrer level to account for the potential correlation within these referral decisions.

@@

Second, the independence of irrelevant alternatives. This assumption could be violated if peers within the same SES group are viewed as close substitutes, where adding similar alternatives distorts choice probabilities. While this concern may have some validity in our setting,\footnote{Among participants making referrals to two different individuals, half refer to someone else from the same SES, suggesting potential substitutability within SES groups.} Alternative discrete choice models that relax IIA are computationally prohibitive given our large dataset.\footnote{Models such as nested logit become computationally intractable with over 250,000 observations across 734 individuals.} We therefore proceed with the conditional logit framework while acknowledging this limitation.

\section{Design} \label{Design}
We designed an online experiment to assess peer referral selection from an SES perspective and to evaluate the causal effect of providing a bonus to referral candidates. The experimental design consisted of two incentivized tasks administered in the following sequence: First, participants completed belief elicitation tasks about their own performance on the national university entry exam. Second, they completed the main referral task, nominating peers based on exam performance in two academic areas. Finally, participants reported beliefs about their referrals' performance and provided demographic information. This structure allowed us to measure both the accuracy of participants' beliefs and their referral behavior under controlled incentive conditions. Figure \ref{timeline} shows the experimental timeline, and detailed instructions are provided in Appendix \ref{instructions}.

@@

\subsection{Performance measures}
To establish an objective basis for referral performance, we use national university entry exam scores (SABER 11). All Colombian high school students take the SABER 11 exam at the end of their final year as a requirement for university admission. The scores from this exam provide pre-existing, comparable measures of performance. By using existing administrative data, we also ensure that all eligible students have comparable performance measures.

The exam consists of five areas (critical reading, mathematics, natural sciences, social sciences, and English). We focus on critical reading and mathematics as these represent two independent and fundamental skills. Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history). Mathematics assesses students' competency in using undergraduate level mathematical tools (e.g., reasoning in proportions, financial literacy). These together capture performance in comprehending and critically evaluating written material as well as reasoning and problem-solving abilities.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the participants can earn money based on their referral's performance. The \textbf{Bonus} treatment adds a fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

\section{Sample, Incentives, and Procedure} \label{SID}
We invited all 4,417 undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection (19\% response rate). Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together (88\% success rate). We randomly allocated participants to either \textbf{Baseline} or \textbf{Bonus} treatments.

@@

We now examine how connection intensity relates to network size and composition. First, if two students take more courses together, it is very likely that they are in the same academic program. We plot this relationship in Figure \ref{share_same_program_by_tie}: As students take more than 5 courses together, the share of students in their enrollment network from the same academic program quickly exceeds 90\%. Second, because students sort into specialized academic programs, increases in courses taken together should result in decreases in connections. We plot this relationship in Figure \ref{conns_by_tie}: As students take more than 5 courses together, the size of their enrollment network drops dramatically from above 210 to below 50. These patterns reveal that while participants' overall networks are large with relatively few courses taken together on average, they spend most of their time at the university within smaller, more intensive groups of peers from the same academic program.

@@

Participants made one referral for math and one referral for the reading part of the university entry exam from their enrollment networks. We observe 1,342 referrals from 734 participants in our final dataset. More than 90\% of these consist of participants referring for both exam areas (see Appendix Table \ref{tab:referral_distribution}). About 70\% of these referrals go to two separate individuals. We compare the outcomes across exam areas for referrals only going to separate individuals in Appendix Table \ref{tab:referral_by_area} and all referrals in Appendix Table \ref{tab:referral_area}. In both cases, we find no meaningful differences between referrals made for Math or Reading areas of the entry exam. As referrals in both exam areas come from the same enrollment network, we pool referrals per participant and report their averages in our main analysis to avoid inflating statistical power in our comparisons.

@@

First, referrals go to peers with whom the referrer has taken around 14 courses with on average, compared to almost 4 on average with others in their network (see Figure \ref{tie_hist}). This difference of 10.1 courses is significant ($t = 34.98$, $p < 0.001$), indicating that referrers choose individuals with whom they have stronger ties. While the median referral recipient has taken 12 courses together with the referrer, the median network member has shared only 2.8 courses. The interquartile range for referrals spans from 7.5 to 19.5 courses, compared to just 2.1 to 4.0 courses for the broader network, highlighting the concentration of referrals among peers with higher connection intensity. In addition, 93\% of referrals go to students in the same program as the referrer.

@@

Examining patterns by referrer SES reveals larger deviations. Low-SES referrers have the largest same-SES deviation, referring 12.9 percentage points more to low-SES students than their network composition suggests, while high-SES referrers under-refer to low-SES students by 10.9 percentage points (see Figure \ref{fig:all_ses_referral_rates}). These descriptive findings suggest that referral behavior diverges most from underlying network structure when SES differences are most pronounced and motivate our formal analysis.

>>>>

\title{When Proximity Isn't Enough: Network Segregation and SES Bias in Referrals}
% \\ \large Experimental Design \\
% \title{Skills, Inequality, and Referrals}
\author{Manuel Munoz\footremember{liser}{Luxembourg Institute of Socio-Economic Research}, Ernesto Reuben\footnote{Division of Social Science, New York University Abu Dhabi} \footrecall{liser}, Reha Tuncer\footnote{University of Luxembourg}}  
% \footnote{Center for Behavioral Institutional Design at New York University Abu Dhabi} 
\begin{document}
\linenumbers 
% HERE FOR LINES
\maketitle

\section*{Abstract}
The share of high-SES connections in one's network is a strong correlate of labor market income. We investigate whether SES biases in referral selection exacerbate differences high-SES connection shares. We conduct a lab-in-the-field experiment with 734 Colombian university students who make incentivized referrals from their enrollment networks. Randomizing participants between performance-only incentives and performance plus a fixed bonus for referral recipients, we find that referrals go to high-performing peers with whom they take many courses together, regardless of incentives. While low-SES referrers exhibit strong in-group preferences, middle- and high-SES referrers show no bias toward other groups and refer along the network shares of each SES groups. We find that network segregation, driven by program selection based on SES, limits cross-SES referral opportunities for even without an explicit SES bias. These suggest institutional policies promoting cross-SES contact are key for reducing SES-based inequalities.
\medskip \\
\textbf{JEL Classification:} C93, J71, D85, Z13 \\
\textbf{Keywords:} inequality, economic mobility, peer networks, class discrimination, homophily

\pagebreak
\section{Introduction}
Equally qualified individuals face different labor market outcomes based on their socioeconomic status \citep{stansbury2024class}. This persistent inequality undermines meritocratic ideals and represents a substantial barrier to economic mobility. A key driver of SES-based inequality in the labor market stems from differences in social capital.\footnote{See for example \citet{loury_dynamic_1977,bourdieu_forms_1986} for pioneering work on the relationship between social position and human capital acquisition.} Economic connectivity, defined as the share of high-SES connections among low-SES individuals, is an important facet of social capital because it correlates strongly with labor market income \citep{chetty2022na}. In this sense, a lack of social capital means lack of access to individuals with influential (higher paid) jobs and job opportunities. It implies having worse outcomes when using one's network to find jobs conditional on the capacity to leverage one's social network.\footnote{See for example \citet{lin1981asr,mouw2003asr} for differential outcomes while using contacts in job search, and \citet{smith2005ajs,Pedulla2019RaceAN} specifically for the effects of race conditional on network use.}

Research on economic connectivity has focused on two distinct mechanisms that shape cross-SES connections: network composition (who you have the chance to meet inside an institutional environment) versus individual preference (who you choose to connect with among those available). The prevailing hypothesis emerging from the seminal work of \citet{chetty2022n} is that increasing exposure to high-SES individuals will lead low-SES individuals to connect with them at higher rates. Universities, in this regard, represent a particularly promising setting as they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, credible evidence on biases in individual preferences to connect across SES groups remains limited. One important reason for this gap is the challenge of creating controlled environments that isolate SES biases while accounting for natural variations in network compositions.

We overcome this challenge through a lab-in-the-field experiment at a Colombian university. Focusing on the role of SES in referral selection, we studied whether individuals who were asked to refer a peer tended to refer a same-SES candidate. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam, and to incentivize performance-based referral selection, participants earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals where SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course the referrer had taken with all other undergraduate students at the university (more than 4,500 individuals). It allowed us to observe both characteristics of every potential referral candidate, and the intensity of interaction between the two, which we measured by the number of courses taken together. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. We designed the \textbf{Bonus} condition to make SES biases in referral selection even more salient. The fixed bonus created incentives to refer peers even if they performed less well, potentially amplifying the relevance of other factors like the SES bias and the connection intensity.

We find that referrals consistently go to higher-performing peers with high connection intensity (14 vs. 4 courses), regardless of the conditions and the exam area. Pooling across these, we find that SES bias in referral selection is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no bias toward other groups.

With 93\% of referrals going to peers within the same academic program with whom referrers have taken many courses together, we find that network composition rather than SES biases better explain the observed referral patterns. At the connection intensity where referrals typically occur (median 12 courses together), network segregation becomes stark: low-SES networks contain 44.5\% low-SES peers versus 35\% university-wide (27\% increase), while high-SES networks contain only 15.7\% low-SES peers (55\% decrease from the university average). This segregation means that even without any bias against low-SES peers, high-SES referrers rarely encounter low-SES candidates in their practical choice sets.

Looking for potential mechanisms driving the segregation in enrollment networks, we identify program selection as key. Program fees at our partner university are fixed on a cost basis, and less than 5\% of undergraduates qualify for scholarships. One consequence of these policies is that SES groups end up sorting into programs on the basis of their costs, where some programs cost up to six times more on a yearly basis. To sum, even though low-SES are exposed to higher-than-population shares of high-SES students, and high-SES are not biased toward other SES groups, meaningful interaction opportunities at the university are genuinely limited.

Our findings should be interpreted with some scope conditions. First, our referrals have no direct job consequences, and participants refer under anonymity. These may represent a lower stake environment for referrers with no potential reputational concerns. Nevertheless, we replicate typical findings from earlier referral experiments where performance-based incentives brings in qualified candidates from participant networks (e.g., \citet{witte2021workers,beaman_who_2012}).

Second, enrollment networks capture classroom-based interactions and their intensity rather than broader networks of close friendships. While our approach has clear advantages over self-reported friendship network elicitation which suffers from censoring due to limitations in size \citep{griffith2022jle}, triangulating it with an additional method (e.g., social media friendship data) could provide useful for better identifying actual interactions at the university. Still, we find that connection intensity predicts referral selection well beyond same program affiliation, suggesting it does capture meaningful variation in social interactions in some dimension.

Finally, our setting examines SES bias within a single institution where cross-SES contact is possible, and the networks of different SES groups are separated due to program selection. The generalizability to contexts with different institutional structures remains an open question for future research.

@@

A subset of the literature focuses on SES-based differences in social capital and network formation \citep{chetty2022na,engzell2025ajos,michelman2022tqjoe}, with connection intensity \citep{gee2017jole,kramarz2014res,wang2013aejae,sterling2014ms} and homophily \citep{mcpherson2001birds,currarini2009e,jackson2022,bolte2024jle,montgomery_social_1991} driving differences across groups. Based on the pioneering work of \citet{currarini2010pnas}, we contribute by identifying two different types of homophily, and separate whether differential outcomes stem from network composition (who you know) versus taste-based biases (who you choose to interact with). Our findings suggest that rather than focusing on taste-based biases,  implementing mixed-program courses to increase across-SES connection intensity should be a clear policy goal in order to reduce SES-based network segregation (see e.g., for an institutional intervention by \citet{rohrer2021po}).

Methodologically, we contribute to the literature on job referral experiments. This literature provides causal evidence on why referrals in the labor market are prevalent,\footnote{Referrals solve frictions in the search and matching process and benefit both job-seekers and employers \citep{topa2019social}. Referral candidates tend to get hired more often, have lower turnover, and earn higher wages \citep{obukhova2013ms,brown2016informal,dustmann2016referral}.} finding that performance-based incentives bring in qualified candidates otherwise not identified by demographic characteristics \citep{beaman_who_2012,pallais_why_2016,friebel2023employee,witte2021workers}, and the consequences of relying upon referral hiring, which come at the cost of disadvantaging certain groups \citep{beaman_job_2018,hederos2025gender}. We extend this literature by causally evaluate the effect of a sizeable monetary bonus for the referral candidate and exploring SES biases in referral selection.

The remainder of the paper is organized as follows. Section \ref{BS} begins with the background and setting in Colombia. Section \ref{ES} presents the empirical strategy and Section \ref{Design} presents the design of the experiment. In Section \ref{SID} we describe the experimental sample, incentives and the procedure. Section \ref{Results} discusses the results of the experiment and Section \ref{MandRC} discusses potential mechanisms and robustness checks. Section \ref{Conclusion} concludes. The Appendix presents additional tables and figures as well as the experiment instructions.

@@

\subsection{Inequality and SES in Colombia}
Our experiment took place in Colombia, a country that consistently ranks highly in terms of economic inequality. The richest decile of Colombians earn 50 times more than the poorest decile \citep{barretoherrera2024regional,comisioneconomicaparaamericalatinayel2023}. This economic disparity creates profound differences in outcomes across SES groups in terms of education, geographic residence, language, manners, and social networks \citep{garciavillegas2021dimension, angulo2012movilidad,garcia2015loteria}. While similar patterns also exist elsewhere, Colombia's pronounced economic inequality makes educational and cultural differences across SES groups particularly visible.

@@

We rely on Colombia's established estrato classification system to measure SES in our study. In 1994, Colombia introduced a nationwide system that divides the population into six strata based on  ``similar social and economic characteristics'' \citep[p. 102]{hudson_colombia_2010}. Designed for utility subsidies from higher strata to support lower strata, the system aligns with and reinforces existing social class divisions \citep{uribe2008estratificacion, guevara2019spatializing}. It is also widely used by policymakers and in official statistics \citep{fergusson2021desigualdad} and well known to by the public. Using the estrato system, we categorize students in strata 1-2 as low-SES, strata 3-4 as middle-SES, and strata 5-6 as high-SES.

\begin{figure}[H]
  \centering
  \caption{Income, performance, and university choice in Colombia}\label{school_quality_income}
  \includegraphics[width=0.6\linewidth]{../figures/school_quality_income.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure shows the average score national university entry exam by monthly family income and type of higher education institution. With average student scores in the 65-70 band, the private university where we conducted this study caters to both low- and high-income students. Figure reproduced from \citet{fergusson2021distincion}.
  \end{tablenotes}
\end{figure}

@@

Undergraduate students at the university choose among 32 different academic programs. Students take between 5 and 7 courses per semester, and programs last anywhere between 4 and 12 semesters (2 to 6 years). The majority (64\%) of students are enrolled in the 10 programs described in Appendix Figure \ref{program_share}. Medicine, the largest program by size at the university, lasts for 12 semesters, followed by engineering programs at 10 semesters. Most remaining programs last for about 8 to 10 semesters, with specialized programs for immediate entry into the workforce lasting only 4 semesters. Academic program choice thus shapes students' connections at the university, influencing both who they encounter in classes and the frequency of these interactions.

To map these social connections, we construct enrollment networks using administrative data. For each participant, we identify all other undergraduate students with whom they have taken at least one course and create their individual network of university connections. The size of this network depends on how many students a participant has encountered through coursework, while the intensity of connection is measured by the number of courses taken together. This approach provides a complete picture of each participant's social environment at the university, and includes detailed characteristics (i.e., SES, academic program, performance) for both the participant and every person in their network.

@@

where $Y_{ij} = 1$ if referrer $i$ chooses referral candidate $j$, and 0 otherwise. We set middle-SES as the base category, so $\beta_1$ is the log-odds estimate for referring low- and high-SES candidates relative to middle-SES. $X_{ij}$ includes the remaining characteristics of referral candidates in the enrollment network that improve model fit such as entry exam scores and the number of courses taken together with the referrer. These continuous variables are standardized using means and standard deviations calculated by first computing network-level statistics for each referrer, then averaging across all 734 networks.\footnote{Each referral candidate's entry exam score and the number of courses they have taken with the referrer is standardized using these sample-level statistics. The standardization formula is $z_{i} = (x_{i} - \bar{X}) / \sigma$, where $\bar{X}$ and $\sigma$ are the average mean and standard deviation across participant networks for the measure.} The individual fixed effects $\alpha_i$ control for referrer-specific factors that might influence both network formation and referral decisions. Because we observe two referrals (one per exam area) from each referrer, we cluster standard errors at the referrer level and account for the potential correlation in the error terms.

@@

Second, the independence of irrelevant alternatives. This assumption could be violated if peers within the same SES group are viewed as close substitutes, where adding similar alternatives distorts choice probabilities. While this concern may have some validity in our setting,\footnote{Among participants making referrals to two different individuals, half refer to someone else from the same SES, suggesting potential substitutability within SES groups.} alternative discrete choice models that relax IIA are computationally prohibitive given our large dataset.\footnote{Models such as nested logit become computationally intractable with over 250,000 observations across 734 individuals.} We therefore proceed with the conditional logit framework while acknowledging its limitations.

\section{Design} \label{Design}
We designed an experiment to assess SES biases in referral selection and to evaluate the causal effect of providing a bonus to referral candidates. The experimental design consisted of two incentivized tasks administered in the following sequence: First, participants completed belief elicitation tasks about their own performance on the national university entry exam. Second, they completed the main referral task, nominating peers based on exam performance in two academic areas. Finally, participants reported beliefs about their referrals' performance and provided demographic information. This structure allowed us to measure the accuracy of participants' beliefs and observe their referral decisions in a controlled setting. Figure \ref{timeline} shows the experimental timeline, and detailed instructions are provided in Appendix \ref{instructions}.

@@

\subsection{Performance measures}
To establish an objective basis for referral performance, we use national university entry exam scores (SABER 11). All Colombian high school students take the SABER 11 exam at the end of their final year as a requirement for university admission. The scores from this exam provide pre-existing, comparable measures of performance.

The exam consists of five areas (critical reading, mathematics, natural sciences, social sciences, and English). We focus on critical reading and mathematics as these represent independent and overarching skills. Critical reading evaluates competencies necessary to understand, interpret, and evaluate texts found in everyday life and broad academic fields (e.g., history). Mathematics assesses students' competency in using high school level mathematical tools (e.g., reasoning in proportions, financial literacy). These together capture performance in comprehending and critically evaluating written material as well as reasoning and problem-solving abilities.

@@

\subsection{Bonus Treatment}
To examine how different incentive structures affect referral selection, we randomly assign a fixed bonus payment for students who get a referral. In the \textbf{Baseline} treatment, only the participants, i.e., those who make referrals, can earn money based on their referral's performance. The \textbf{Bonus} treatment adds a fixed payment of \$25 to the peer who gets the referral. This payment is independent of the referral's actual performance (see Table \ref{tab:treatments}).

@@

\section{Sample, Incentives, and Procedure} \label{SID}
We invited all 4,417 undergraduate students who had completed their first year at the university at the time of recruitment to participate in our experiment. A total of 837 students participated in the data collection (19\% response rate). Our final sample consists of 734 individuals who referred peers with whom they had taken at least one class together (88\% success rate).

@@

We now examine how connection intensity relates to network size and composition. First, if two students take more courses together, it is very likely that they are in the same academic program. We plot this relationship in Figure \ref{share_same_program_by_tie}: As students take more than 5 courses together, the share of students in their enrollment network from the same academic program quickly exceeds 90\%. Second, because students sort into specialized academic programs, increases in courses taken together should result in decreases in connections. We plot this relationship in Figure \ref{conns_by_tie}: As students take more than 5 courses together, the size of their enrollment network drops dramatically from above 210 to below 50. These patterns reveal that while participants' overall networks are large with relatively few courses taken together on average, they are more frequently in contact within a much smaller group of peers from the same academic program.

@@

Participants made one referral for math and one referral for the reading part of the university entry exam from their enrollment networks. We observe 1,342 referrals from 734 participants in our final dataset. More than 90\% of these consist of participants referring for both exam areas (see Appendix Table \ref{tab:referral_distribution}). About 70\% of these referrals go to two separate individuals. We compare the outcomes across exam areas for referrals only going to separate individuals in Appendix Table \ref{tab:referral_by_area} and all referrals in Appendix Table \ref{tab:referral_area}. In both cases, we find no meaningful differences between referrals made for math or critical reading areas of the entry exam. As referrals in both exam areas come from the same enrollment network, we group referrals per participant and report average outcomes.

@@

First, referrals go to peers with whom the referrer has taken around 14 courses with on average, compared to almost 4 on average with others in their network (see Figure \ref{tie_hist}). This difference of 10.1 courses is significant ($t = 34.98$, $p < 0.001$), indicating that referrers choose individuals with whom they have higher contact intensity. While the median referral recipient has taken 12 courses together with the referrer, the median network member has shared only 2.8 courses. The interquartile range for referrals spans from 7.5 to 19.5 courses, compared to just 2.1 to 4.0 courses for the broader network, highlighting the concentration of referrals among peers with higher connection intensity. In addition, 93\% of referrals go to students in the same program as the referrer.

@@

Examining patterns by referrer SES reveals larger deviations. Low-SES referrers have the largest same-SES deviation, referring 12.9 percentage points more to low-SES students than their network composition suggests, while high-SES referrers under-refer to low-SES students by 10.9 percentage points (see Figure \ref{fig:all_ses_referral_rates}). These descriptive findings suggest that referral selection in SES terms diverges most from underlying network structure when SES groups are further apart, and motivate our formal analysis.
